---
title: "Heatmaps & Clustering"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, include = T)
```

## Shiny Presentation

This R Markdown presentation is made interactive using Shiny. The viewers of the presentation can change the assumptions underlying what's presented and see the results immediately. 

To learn more, see [Interactive Documents](http://rmarkdown.rstudio.com/authoring_shiny.html).

## Distance

In general, a clustering algorithm groups together proximal data points in a parameter space.
While there are several distance measures, Euclidean distance is the most common and most intuitive. Each parameter is treated as an axis in a coordinate system, and distances can be found trigonometrically. 

## Cell positions



## Hierarchical Clustering

Each point is joined to it's nearest neighbor in a dendrogram.

```{r}
# read data
# ATACseq counts
counts <- read.table('counts.tsv')

# compute distances
distance <- dist(t(as.matrix(counts)))

# create dendrogram
hclusts <- hclust(distance)

# plot dendrogram
# pdf() opens a connection to an image file & writes any subsequent plots to the file
pdf('dendrogram.pdf')

plot(hclusts)

# dev.off() closes the connection. Otherwise each subsequent plot will overwrite the previous one.
dev.off()

# cut dendrogram into 5 most distal clusters
cutree(hclusts,5)

```
## Kmeans clustering

The number of clusters is predefined. K rows are selected as centers, other rows are iteratively assigned to a center to minimize distance.

```{r}
km <- kmeans(distance,5)

```

## ComplexHeatmap

Both these methods can be used in heatmaps.

```{r}
library(ComplexHeatmap)
hm <- Heatmap(counts)

pdf('counts.pdf')
draw(hm)
dev.off()

```

## Z-score normalization
Counts can be converted to z-scores, which can be thought of as standard deviations from the mean.
When applied over rows, it gives the deviation from the average value of a feature. This is a way of normalizing dissimilar *measurements*, e.g. characteristics of embryos.
When applied over columns, it gives the deviation from the average expression in a replicate. This normalizes for dissimilar *experiments* e.g. differences in sequencing depth.
```{r}
zscore <- function(x) {
	m <- mean(x,na.rm=T)
	x <- (x-m)/sd(x,na.rm=T)
	return(x)
}

# apply() accepts a matrix & passes each row or column independently to a specified function
# the second argument is the margin (1 for rows, 2 for columns)
z.rows <- apply(counts,1,zscore)
z.cols <- apply(counts,2,zscore)

pdf('counts_z.pdf')
draw(Heatmap(z.cols,show_row_names = F))
dev.off()

```

## color scale

By default, the color scale includes the entire range of the data, but it can be adjusted to fit only the 0.01-0.99 quantiles. 
This requires defining a color scale with the colorRamp2 library.
```{r}
col.z <- function(x,quant=.01) {
	# require() loads a library within a function call.
	require(circlize)
	# colorRamp2() accepts a numeric vector and a color vector.
	# It returns a function that assigns color values to numeric values.
	colorRamp2(
	      c(quantile(x,quant,na.rm=T),0,quantile(x,1-quant,na.rm=T)),
	      c('blue','white','red')
	)
}

hm.cols <- col.z(z.cols)

hm <- Heatmap(z.cols,show_row_names = F,
	      col=hm.cols)

# note that hm.cols() is a *function* and can be invoked independently of Heatmap()
hm.cols(-3:3)

draw(hm)

# divide heatmap using k-means clustering
hm.k <- Heatmap(z.cols,show_row_names = F, km=5,
	      col=hm.cols)

draw(hm.k)

# the same can be done for columns
hm.k <- Heatmap(z.cols,show_row_names = F, km=5,column_km=5,
	      col=hm.cols)

pdf('counts_z_kmeans.pdf')
draw(hm.k)
dev.off()
```
## split

A heatmap can also be split into specified groups.
```{r}
# experimental design
design <- read.delim('ataclib.txt',row.names=1)
hm.k <- Heatmap(z.cols,show_row_names = F, km=5,
	      col=hm.cols,
	      column_split = design$condition,column_title_rot = 90)

pdf('counts_z_kmeans.pdf')
draw(hm.k)
dev.off()
```

## simpleAnnotation

ComplexHeatmap allows adding HeatmapAnnotations. This allows us to show multiple data characteristics.
```{r}
ann <- columnAnnotation(condition=design$condition,time=design$time)

hm.k <- Heatmap(z.cols,show_row_names = F, km=5,
	      col=hm.cols,top_annotation = ann)
#The plot size can be adjusted by specifying height & width
pdf('counts_z_kmeans.pdf',height=20,width=20)
draw(hm.k)
dev.off()
```

## setting heatmap element size

```{r}

hm.k <- Heatmap(z.cols,show_row_names = F, km=5,
	      col=hm.cols,top_annotation = ann,
	      heatmap_width = unit(2,'npc'),
	      heatmap_height = unit(2,'npc'),
	      width = unit(25.4,'cm'),height=unit(25.4,'cm'))

draw(hm.k)

hm.cell <- function(
		x,...,
		cell.h=NULL,cell.w=NULL,
		height=NULL,width=NULL,
		heatmap_height=NULL, heatmap_width=NULL,
		units='in'
){
	require(ComplexHeatmap)
	if(!is.null(cell.h)) height <- unit(nrow(x)*cell.h,units)
	if(!is.null(cell.w)) width <- unit(ncol(x)*cell.w,units)
	return(Heatmap(x,...,height=height,width=width))
}

hm.k <- hm.cell(z.cols,show_row_names = F, km=5,
	      col=hm.cols,top_annotation = ann,
	      heatmap_width = unit(2,'npc'),
	      heatmap_height = unit(2,'npc'),
	      cell.h=.01,cell.w=.2,units='in')

pdf('counts_z_kmeans.pdf',height=20,width=20)
draw(hm.k)
dev.off()

```

## More complex annotations

```{r}
get.cols <- function(x,...){
  setNames(rainbow(length(unique(x)),...),unique(x))
}

better.color <- function(x,...){
  anno_simple(x,col=get.cols(x,...))
}

ann <- columnAnnotation(
  condition=better.color(design$condition),
  time=better.color(design$time,s=.5),
  ncells=anno_barplot(design$ncells,height = unit(1,'in'))
)

row.ann <- rowAnnotation(
  count.value=anno_lines(
    cbind(mean=apply(counts,1,mean),
          SD=apply(counts,1,sd)),
    width = unit(1,'in'),
    gp = gpar(col=c(1,2))
  )
)

hm.k <- hm.cell(z.cols,show_row_names = F, km=5,
	      col=hm.cols,top_annotation = ann,
	      right_annotation=row.ann,
	      heatmap_width = unit(2,'npc'),
	      heatmap_height = unit(2,'npc'),
	      cell.h=.01,cell.w=.2,units='in',
	      name='z-score')

pdf('counts_z_kmeans.pdf',height=20,width=20)
draw(hm.k)
dev.off()
```

#Adding custom legends

```{r}
lgd <- list(
  Legend(unique(design$condition),
         legend_gp=gpar(fill=get.cols(design$condition)), 
         title = 'condition'),
  Legend(unique(design$time),
         legend_gp=gpar(fill=get.cols(design$time,s=.5)),
         title = 'time'),
  Legend(c("mean","SD"),
         type='lines',
         legend_gp = gpar(col=c(1,2)),
         title='count value')
)

pdf('counts_z_kmeans.pdf',height=20,width=20)
draw(hm.k, annotation_legend_list=lgd)
dev.off()
```